\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-08 paper.tex
Various edits from Yan and Ana

% FIXME abstract needs some rewriting, we're placing too much emphasis
% on the coalescent there.
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require complete precision about all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occurred,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes where they appear unary in local trees.
}
\end{figure}

@@

Uncertainty is a central element of inference about recombination,
and, even with small samples, we can only expect to uniquely identify
lineages and breakpoints in special cases.
% Keeping comments here - otherwise this para is clearly missing stuff
[summarise info inferable about recomb in few sentenances here?
Cite: \citep{myers2002detection,hayman2023recoverability}]
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
The fact that the classical ``event'' ARG (eARG) encoding
\emph{requires} completely precise estimates about recombination
is therefore a fundamental limitation, and a central message
of this paper.

@@

% Optimisation problems to do with ARGs and other phylogenetic networks
There has also been substantial interest in formulating and answering
fundamental questions about properties
of the ARG as a discrete graph structure, focussing on the ARG topology without considering
either branch lengths or indeed the generating process.
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results that largely do not
have direct relevance to the inference problems described above.

@@

% FIXME putting this here for now just to see if this will work.

% \section*{Event ARGs}\label{eARG}
% In keeping with its original derivation in terms of a stochastic
% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
% ARG data structure is
% in terms of events.
% Nodes represent
% common ancestor or recombination events that occurred in the
% history of some set of sample genomes, and edges represent ancestral
% lineages~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage, and in a recombination
% event a lineage is split into two independent
% ancestral lineages.
% % The final vital
% % detail is that we associate a breakpoint with each recombination
% % event.
% % This information is sufficient to uniquely
% % define the local genealogical trees at every position along the genome,
% % which is the basic requirement for a data structure encoding
% % genome-wide genetic ancestry.
% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
% is usually concretely encoded by associating a recombination breakpoint
% (which is null for common ancestor events) and an ordered list of parents
% with each node. We refer to such a  data structure as an
% ``event ARG'' (eARG).
% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.

% This ordering requirement, while straightforward
% to describe, has some practical drawbacks. For example,
% % Surely the meaning is clear from the context here and we don't need
% % qualify the meaning of ``ARG''
% when using this representation and simulating an ARG backwards in time,
% the first event older than a recombination may involve the lineage carrying the
% ancestry to the right of the breakpoint rather
% than the left, and so we cannot emit edges as they are generated.
% Such issues can be worked around, of course, but depending on the ordering of
% otherwise indistinguishable objects is generally problematic.

% \begin{figure}
% \centering
% % \begin{tabular}{cc}
% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
% \tikzset{greynode/.style={circle,fill,inner sep=1},
% nodelabel/.style={font=\footnotesize}}

% \node (s0) [greynode] at (0, 0) {};
% \node (s1) [greynode] at (3, 0) {};
% \node (s2) [greynode] at (6, 0) {};
% \node (s3) [greynode] at (3, 1) {};
% \node (s4) [greynode] at (1, 2) {};
% \node (s5) [greynode] at (5, 3) {};
% \node (s6) [greynode] at (3, 4) {};

% % \node [anchor=north west] at (0,6) {A};
% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

% %% Edges
% \draw (s1) -- (s3);
% \draw (s0) |- (s4);
% \draw (s4) -- (2,2) |- (s3);
% \draw (s4) |- (s6);
% \draw (s3) -- (4,1) |- (s5);
% \draw (s2) |- (s5);
% \draw (s5) |- (s6);

% % \node [anchor=north west] at (9,6) {B};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{c|c|l}
% % \multicolumn{2}{c}{Breakpoints}\\
% Node & Breakpoint & Parents\\
% \hline
% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{g}$ & $\varnothing$ & []\\
% \end{tabular}};

% % \node [anchor=north west] at (9,6) {C};
% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
% % \begin{tabular}{ll}
% % Edges & Intervals\\
% % \hline
% % $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
% % $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
% % $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
% % $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
% % $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
% % $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
% % $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
% % \end{tabular}};

% \end{tikzpicture}
% \caption{\label{fig-arg-data-structure}
% A graphical depiction of a classical event ARG and a concrete
% encoding of the information. Node \noderef{d} is a recombination
% event as it is associated with a non-null breakpoint, and has
% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
% common ancestor events.
% }
% \end{figure}

% % What do we mean by an eARG, formally?
% To be concrete, we can formally define the
% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
% is an ordered list of edges defining a directed acyclic graph and
% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
% is a function mapping nodes to recombination breakpoints,
% such that $\sigma(u) = x$
% if $u$ is a recombination event with breakpoint $x$ and
% $\sigma(u) = \varnothing$ otherwise.
% (This is equivalent to having different ``types''
% of node, where breakpoints are associated with recombination
% nodes only).
% Each edge $e_j = (c, p)$ describes an edge between
% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% % For each recombination node $u$ we must
% % have exactly two edges in $e$ such that $u$ is the child.
% This encoding of an eARG is equivalent to the output
% of programs such as
% \texttt{ARGweaver}~\citep{rasmussen2014genome}
% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
% Note that the description above captures only the
% graph topology: if we also wish to know the branch lengths we need
% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
% which defines the time of each node.

\section{The Big and Little ARG}
\label{app-big-and-little-arg}

Here we review the two predominant stochastic processes which construct ARGs:
the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
 \cite{hudson1983properties}. The big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

>>>>

% FIXME abstract needs some rewriting, we're placing too much emphasis
% on the coalescent there.
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on specific details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require complete precision about all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

% NOTE: important to emphasise TWO things here, events and vagueness
% actual encoding mechanism (i.e. how trees fit into the ARG)
A common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occurred,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes where they are unary in local trees.
}
\end{figure}

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from sequencing data. For ARGs under 
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
positions can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. For general
ARGs, for instance those representing pedigrees, or genealogies of organisms with
low mutation rates and complex patterns of recombination, such inference can be
even more challenging. The fact that the eARG encoding \emph{requires} completely
precise estimates about recombination is therefore a fundamental limitation, which
is a central message of this paper.
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121

@@

% Optimisation problems to do with ARGs and other phylogenetic networks
There has also been substantial interest in formulating and answering
fundamental questions about properties
of the ARG as a discrete graph structure, focusing on the ARG topology without considering
either branch lengths or indeed the generating process.
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results~\citep{gusfield2014recombinatorics}
that are often not of
direct relevance to the inference problems described above.

@@

\section{The Big and Little ARG}
\label{app-big-and-little-arg}
Here we review the two predominant stochastic processes which construct ARGs:
the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
 \cite{hudson1983properties}. The big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.
